---
title: "CDC Mortality in USA Between 1999-2010"
subtitle: DATA 608 Module 3 
author: "Gehad Gad"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    toc: true
    toc_float: true
  pdf_document: default
---

```{r, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
#Libraries required
library(ggplot2)
library(plotly)
library(shiny)
library(dplyr)
library(rsconnect)
```



```{r}
#Read the data
df <- read.csv("https://github.com/charleyferrari/CUNY_DATA_608/raw/master/module3/data/cleaned-cdc-mortality-1999-2010-2.csv")
```


```{r}
#Display the first few rows in the data
head(df)
```

```{r}
#Number of colums
ncol(df)
#Names of columns
colnames(df)
```

**There are 6 variables in this dataset(ICD.Chapter, State, Year, Deaths, Population, Crude.Rate) and 9961 observations.** 


# Question 1: 

As a researcher, you frequently compare mortality rates from particular causes across different States. You need a visualization that will let you see (for 2010 only) the crude
mortality rate, across all States, from one cause (for example, Neoplasms, which are effectively cancers). Create a visualization that allows you to rank States by crude mortality for each cause of death.


```{r}
# Subset for 2010 only
df_2010 <- subset(df, Year==2010)
```

```{r}
# Use plotly library to see the Crude rate for 2010 year for Mental and behavioural disorders
Data <- df %>%
  filter(., Year == "2010" & ICD.Chapter == "Mental and behavioural disorders") %>% 
  arrange(order(State), Crude.Rate)                   


head(Data)
```

```{r}

MBD2 <- df_2010 %>%
  plot_ly(x = ~df_2010$Crude.Rate, y = df_2010$State, type="bar", orientation="h") %>%
  layout(
          title= list(text=paste0("Mental and behavioural disorders","\ncause-based 2010 State Crude Mortality Rates"),font=list(size = 10)),
    xaxis=list(title="Crude Rate"),
    yaxis=list(title="States",
    categoryorder = "array",
    categoryarray = (~State)))

subplot(MBD2)   
```




```{r}
# Get the unique causes of death for 2010
Total_causes1 <- unique(df_2010$ICD.Chapter)
length(Total_causes1)
#Total_causes1 <- as.data.frame(unique(df_2010$ICD.Chapter))
```
**There are 17 diseases (cause of death) in the dataset**



```{r}
States <- unique(df_2010$State)
length(States)
```
**There are 51 state in the dataset**


```{r}
# Create server logic
ui1 <- fluidPage(
  titlePanel("CDC Mortality Rates for 2010"),
  fluidRow(selectInput("cause", "Cause of Death:", choices=sort(Total_causes1))), 
  plotOutput("plot1", height = 800)
)
```


```{r}
# Create user interface definition
server1 <- function(input, output) {
  showrates <- reactive({showrates <- subset(df_2010, ICD.Chapter==input$cause)})
  output$plot1 <- renderPlot({
    ggplot(showrates(), aes(x=Crude.Rate, y=reorder(State, -Crude.Rate)))+
      scale_x_continuous(limits=c(0, max(showrates()$Crude.Rate))+5, expand = c(0,0))+
      geom_segment(aes(yend=State), xend=0, color="blue")+
      geom_point(size=3, color = "red") +
      theme_bw()+
      theme(panel.grid.major.y = element_blank(), axis.title=element_text(size=14))+
      xlab("CDC Mortality Rate for 2010 year") +
      ylab("State") +
      ggtitle(input$cause)
  }) 
}

```


```{r}
# Run the application 
shinyApp(ui = ui1, server = server1)
```

**ShinyApp Link: https://g-gad.shinyapps.io/CDC-Morality/?_ga=2.173588579.1949036595.1633893772-1924490615.1633893772**


# Question 2:

Often you are asked whether particular States are improving their mortality rates (per cause)faster than, or slower than, the national average. Create a visualization that lets your clients
see this for themselves for one cause of death at the time. Keep in mind that the national
average should be weighted by the national population.




```{r}
# Create server logic
ui <- fluidPage(
    titlePanel("Crude Mortality Rate Across All States Vs National Average"), 
    sidebarPanel(
        selectInput("select1", label = strong("State"), 
                    choices = levels(as.factor(df$State)), 
                    selected = 1),
    
        selectInput("select2", label = strong("Cause of Death"), 
                    choices = levels(as.factor(df$ICD.Chapter)), 
                    selected = 1),width = "auto"),
        
        mainPanel(
            plotOutput("distPlot")
        )
    )

```

```{r}
# Create user interface definition

server <- function(input, output) {
    output$distPlot <- renderPlot({
        df %>% 
            group_by(Year, ICD.Chapter) %>%
            mutate(N_Population = sum(Population),
                   N_Count = sum(Deaths), 
                   N_Crude_Rate = 10^5*(N_Count/N_Population)) %>% 
            group_by(Year, ICD.Chapter, State) %>%
            mutate(S_Count=sum(Deaths),
                   S_Crude_Rate=10^5*(S_Count/Population)) %>%
            select(ICD.Chapter, State, Year, N_Crude_Rate, S_Crude_Rate) %>% 
            filter(ICD.Chapter == input$select2, State == input$select1) %>% 
            ggplot() +
            geom_bar(aes(x = Year, weight = S_Crude_Rate)) +
            labs(x = "State", y = "Crude Mortality Rate") + 
            geom_line(aes(x = Year, y = N_Crude_Rate, linetype = "National Average"), col = "red", lwd = 1) +
            scale_linetype(name = NULL) +
            theme_minimal()
        }
    )
    }



```



```{r}
shinyApp(ui = ui, server = server)
```



**ShinyApp Link: https://g-gad.shinyapps.io/CDC-Mortality/?_ga=2.168864865.1949036595.1633893772-1924490615.1633893772**
